---
title: "Coursera Statistical Inference Project Part 1"
author: "Robert C Phillips"
date: "Sunday, September 21, 2014"
output: pdf_document
---

##Problem Statement

The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also also 1/lambda. 

We will use lambda = 0.2 for all of the simulations. In this simulation, we will investigate the distribution of averages.

```{r}
lambda <- .2; nosim <- 1000; sampleSize <- 40; mu <- 1/lambda; sigma <- 1/lambda;
samples <- matrix(rexp(sampleSize * nosim, lambda), nosim)
```

###Question 1
Show where the distribution is centered at and compare it to the theoretical center of the distribution.

```{r}
samplesMean <- apply(samples, 1, mean)

#compute the mean of sample means, should be close to 1/lambda = 5
mresult <- data.frame(mean(samplesMean), mu)
colnames(mresult) <- c("sample mean","theoretical mean")
mresult
```

###Question 2
Show how variable it is and compare it to the theoretical variance of the distribution.

```{r}
#compute the variance of sample means, should be close to sigma^2/sampleSize
vresult <- data.frame(var(samplesMean), sigma^2/sampleSize)
colnames(vresult) <- c("sample var","theoretical var")
vresult
```

###Question 3
Show that the distribution is approximately normal...normalize then plot.

```{r}
cFun <- function(x,n) sqrt(n) * (mean(x) - mu)/sigma

dat <- data.frame(
   means = c(apply(samples, 1, cFun, sampleSize)),
   size = factor(rep(c(sampleSize), nosim)))
```
```{r, echo=FALSE}
library(ggplot2)
g <- ggplot(dat, aes(x = means, fill = size)) + 
    geom_histogram(alpha = .20, binwidth=.3, colour = "black", aes(y = ..density..)) 
g <- g + stat_function(fun = dnorm, size = 2)
g + facet_grid(. ~ size)
```

###Question 4
Evaluate the coverage of the confidence interval for 1/lambda: $\bar X \pm 1.96 \frac{S}{\sqrt{n}}$

```{r}
cFun2 <- function(x,n) mean(x) + c(-1,1) * 1.96/sqrt(n)*sd(x)
samplesci <- t(apply(samples, 1, cFun2, sampleSize))
total <- sum(apply(samplesci, 1, function(i) if (mu >= i[1] && mu <= i[2]) 1 else 0))

cresult <- data.frame(total/nosim, .95)
colnames(cresult) <- c("sample ci","theoretical ci")
cresult
```
