---
title: "Coursera Statistical Inference Project Part 2"
author: "Robert C Phillips"
date: "Sunday, September 21, 2014"
output: pdf_document
---

##Problem Statement
Analyze the ToothGrowth data in the R datasets package.

###Question 1
Load the ToothGrowth data and perform some basic exploratory data analyses 

```{r, echo=FALSE}
data(ToothGrowth); head(ToothGrowth)
boxplot(len~supp+dose,
        data=ToothGrowth, 
        names=c("OJ .5","VC .5","OJ 1","VC 1","OJ 2","VC 2"),
        ylab="tooth length", 
        xlab="supplement and dose")
```

###Question 2
Provide a basic summary of the data.
```{r}
summary(ToothGrowth[ToothGrowth$supp == "OJ",])
summary(ToothGrowth[ToothGrowth$supp == "VC",])
```

###Question 3
Use confidence intervals and hypothesis tests to compare tooth growth by supp and dose. (Use the techniques from class even if there's other approaches worth considering)
```{r,echo=FALSE}
t1 <- t.test(len~supp, data=ToothGrowth[ToothGrowth["dose"]==.5,])
t2 <- t.test(len~supp, data=ToothGrowth[ToothGrowth["dose"]==1,])
t3 <- t.test(len~supp, data=ToothGrowth[ToothGrowth["dose"]==2,])
dat <- data.frame(matrix(c(t1$conf.int[1:2],t2$conf.int[1:2],t3$conf.int[1:2]),3,2,byrow=TRUE))
colnames(dat) <- c("lower","upper");rownames(dat) <- c(".5","1","2")
dat
```

###Question 4
State your conclusions and the assumptions needed for your conclusions. 

Increased dosages increases tooth length.  A 2 mg dosage yields simliar mean for each type of supplement.